---
title: "Publication ready figures with biogrowth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publication ready figures with biogrowth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction  

The **biogrowth** package for R allows for modelling microbial growth using methods developed in the field of predictive microbiology. This vignette is an addition to the main vignette describing the detailed use of the functions in **biogrowth**. Here, we assume the reader is already familiar with **biogrowth**, so we focus explicitly on creating and saving publication ready figures. 

Once installed, **biogrowth** can be loaded with  

```{r setup}
library(biogrowth)
```

We start again by loading the necessary packages

```{r}
library(tidyverse)
```

Several functions in the package make use of stochastic algorithms. For reproducibility, we will set the seed of R's internal random number generator (PRNG) to some arbitrary value

```{r}
set.seed(1241)
```


**Biogrowth** uses the **Cowplot** package and the **Cowplot** theme as default. Therefore, the vanilla figures produced by **biogrowth** already have a nice clean theme and could be used in a publication as is. However, more control over formatting options can be needed when preparing figures for a specific journal.  

**Cowplot** is a package that works with **gplot2**, and therefore the plots can be manipulated in the same way. Arguments that control the way that data is represented such as:
  
* `Line thickness`
* `Line colour`
* `Line type`

would normally go inside the ggplot call, and can be set by providing additional arguments to the **biogrowth** function. 


We start with a prediction using the modified Gompertz model

```{r}
# Specify the model
my_model <- "modGompertz"

# Define the starting parameters (`mu`, `lambda` and `C`), and the initial count in a list
my_pars <- list(logN0 = 2, C = 6, mu = .2, lambda = 25)

# Define the storage times
my_time <- seq(0, 100, length = 1000)

# Call to predict_isothermal_growth using the arguments defined above
static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

# Plotting
plot(static_prediction) 
```

By setting the optional arguments `line_thickness` and `line_colour` to plot(static_prediction), we can control the colour of the plot. 

```{r}
# Plotting
# Make examples

# plot(static_prediction) + 
#   ylab("bert") +
#   theme(axis.line = element_line(colour = 'black', size = 0.1))
```




```{r}
# Axis line thickness via theme()
plot(static_prediction) + 
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 10)) +
  theme(axis.line = element_line(colour = 'black', size = 0.1)) # Adjust size and colour to taste

```

  

```{r}
P <-ggplot(static_prediction$simulation, aes(x= .data$time, y =.data$logN)) + 
  geom_line(color="red") +
  theme_cowplot()

P +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 10)) +
  theme(axis.line = element_line(colour = 'black', size = 0.4))
```

```{r}
# Defining parameters 
my_model <- "Trilinear"
my_times <- seq(0, 30, length = 100)
n_sims <- 1000

# Calling the function
stoc_growth <- predict_stochastic_growth(my_model, my_times, n_sims,
  mean_logN0 = 0, sd_logN0 = .2,
  mean_sqmu = 2,sd_sqmu = .3,
  mean_sqlambda = 4, sd_sqlambda = .4,
  mean_logNmax = 6, sd_logNmax = .5)

# Plotting
 ggplot(stoc_growth$quantiles, aes(x = .data$time)) +
        geom_ribbon(aes(ymin = .data$q10, ymax = .data$q90), fill = "red",  alpha = .3) +
        geom_ribbon(aes(ymin = .data$q05, ymax = .data$q95), fill = "red",  alpha = .2) +
        geom_line(aes(y = .data$q50)) +
        xlab("Time") +
        ylab("logN") +
        theme_cowplot() 
```






# Advanced plotting  

## Full control  
By accessing the data directly from the S3 object, the experienced user can get full control over plotting:  

```{r}
library(cowplot)

# Vanilla biogrowth theme
static_prediction$simulation %>%
  ggplot(., aes(x= time, y =logN )) + 
  geom_line() +
  theme_cowplot()
```

## Combining plots  
  
In order to compare the outcome of simulations with different parameters we can combine simulations into a single plot
#COMMENT This is not the most elegant way, but maybe more understandable like this?
#Well, you get the idea =P

```{r}
# plotting multiple plots together
my_model <- "modGompertz"

my_pars <- list(logN0 = 2, C = 6, mu = .2, lambda = 25)
my_time <- seq(0, 100, length = 1000)
static_prediction1 <- predict_isothermal_growth(my_model, my_time, my_pars)
P1 <- plot(static_prediction1)

my_pars <- list(logN0 = 2, C = 6, mu = .5, lambda = 10)
my_time <- seq(0, 100, length = 1000)
static_prediction2 <- predict_isothermal_growth(my_model, my_time, my_pars)
P2 <- plot(static_prediction2)

my_pars <- list(logN0 = 2, C = 6, mu = .3, lambda = 15)
my_time <- seq(0, 100, length = 1000)
static_prediction3 <- predict_isothermal_growth(my_model, my_time, my_pars)
P3 <- plot(static_prediction3)


# Plotting using the cowplot plot_grid function
plot_grid(P1, P2, P3, labels = "auto")


```
  
Or, by making a single tibble from the information stored in the static_prediction objects we can make a combined plot, clearly dshowing the differences between  simulations  

```{r}
bind_rows(
  tibble(Plot = rep("P1", 1000), time = static_prediction1$simulation$time, logN = static_prediction1$simulation$logN),
  tibble(Plot = rep("P2", 1000), time = static_prediction2$simulation$time, logN = static_prediction2$simulation$logN),
  tibble(Plot = rep("P3", 1000), time = static_prediction3$simulation$time, logN = static_prediction3$simulation$logN)) %>%
  ggplot(., aes(x = .data$time, y = .data$logN, col = .data$Plot)) + 
  geom_line() +
  theme_cowplot()

```



